---
title: "Homework 3"
author: "Kirya Wagner"
date: "06/01/2024"
format:
  html: 
    toc: true # includes a table of contents in rendered html format
execute: # for all code chunks
    warning: false # do not display warnings
    message: false # do not display messages
---

GitHub Link: https://github.com/kiryawagner/Wagner-Kirya-Homework-03


```{r, echo = FALSE}

# Set up
## reading in packages

# general use
library(tidyverse)
library(readxl)
library(here)
library(janitor)

# visualizing pairs
library(GGally)

# model selection
library(MuMIn)

# model predictions
library(ggeffects)

# model tables
library(gtsummary)
library(flextable)
library(modelsummary)

library(tinytable)

drought_exp <- read_xlsx(path = here("data", 
                                     "Valliere_etal_EcoApps_Data.xlsx"),
                         sheet = "First Harvest")
```

```{r, echo = FALSE}
# cleaning
drought_exp_clean <- drought_exp %>% 
  clean_names() %>% # nicer column names
  mutate(species_name = case_when( # adding column with species scientific names
    species == "ENCCAL" ~ "Encelia californica", # bush sunflower
    species == "ESCCAL" ~ "Eschscholzia californica", # California poppy
    species == "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    species == "GRICAM" ~ "Grindelia camporum", # great valley gumweed
    species == "SALLEU" ~ "Salvia leucophylla", # Purple sage
    species == "STIPUL" ~ "Nasella pulchra", # Purple needlegrass
    species == "LOTSCO" ~ "Acmispon glaber" # deerweed
  )) %>% 
  relocate(species_name, .after = species) %>% # moving species_name column after species
  mutate(water_treatment = case_when( # adding column with full treatment names
    water == "WW" ~ "Well watered",
    water == "DS" ~ "Drought stressed"
  )) %>% 
  relocate(water_treatment, .after = water) # moving water_treatment column after water
```

```{r, echo = FALSE}
## 0. Null model
model0 <- lm(total_g ~ 1, # formula
             data = drought_exp_clean) # data frame

# diagnostics okay 
```

```{r, echo = FALSE}
## 1. total biomass as a function of SLA, water treatment, and species (saturated model)

model1 <- lm(total_g ~ sla + water_treatment + species_name,
             data = drought_exp_clean)

par(mfrow = c(2, 2))

# Diagnostics looks ok!
```

```{r, echo = FALSE}
# 2. total biomass as a function of SLA and water treatment

model2 <- lm(total_g ~ sla + water_treatment,
             data = drought_exp_clean)

# diagnostics look good 
```

```{r, echo = FALSE}
## 3. total biomass as a function of SLA and species

model3 <- lm(total_g ~ sla + species_name,
             data = drought_exp_clean)

# The QQ plot slightly derails from the line at the tails, but the diagnostics overall look okay. 
```

```{r, echo = FALSE}
## 4. total biomass as a function of water treatment and species

model4 <- lm(total_g ~ water_treatment + species_name,
             data = drought_exp_clean)

# diagnostics look okay 
```

```{r, echo = FALSE, results = FALSE}
# Model selection
# compare functions 

model.sel(model0,
          model1, 
          model2, 
          model3,
          model4)
#Out of the four models we have tried, the best model is **model 4, the one that includes biomass as a function water treatment and species**. This is because its delta is 0, and it has the lowest AIC. 
```

```{r, echo = FALSE, results = FALSE}
# Model summary
summary(model4)
```

```{r, echo = FALSE}
# Model predictions 
model_preds <- ggpredict(model4, 
                         terms = c("water_treatment", 
                                   "species_name"))

# use View(model_preds) to see the predictions as a data frame
# use model_preds to see the predictions formatted nicely
```


# Problem 1

## a. Making a table

Table 1: Table displaying model estimates for the null model and models 1-4. Each model displays  predictor estimates in comparison to the reference, and an AIC number.

```{r, results = FALSE}
# comparing models
modelsummary::modelsummary # this function takes a list of models
```

```{r}
models <- list( 
    "null" = model0, # "model name" = model object
    "model 1" = model1,
    "model 2" = model2,
    "model 3" = model3,
    "model 4" = model4)
```

```{r, echo = FALSE}
modelsummary(models, gof_map = c("aic", "r.squared"), stars = TRUE)
    
```


## b. Statistical methods 

In this statistical analysis, I examined the influence of specific leaf area (SLA), species, and water treatment on total growth (g) of the plant by constructing a set of 5 models that compared each variable. These included total biomass as a function of none of the predictors, total biomass as a function of SLA, water treatment and species, total biomass as a function of SLA and water treatment, total biomass as a function of SLA and species, and total biomass as a function of water treatment and species. 

First, I visualized the variables individually against total mass with ggpairs and ggplot, and determined that there seemingly was a correlation between each one. Then I created the models using lm(), and looked at the residuals, distributions, and outliers within their diagnostics to determine if the model was valid.

Next, I compared each model using model.sel(), and determined which one was the best based on their diagnostics and coefficients. The best model not only has a lower AIC and delta of 0, but also visually needs homeostatic residuals, a normal distribution and no influential outliers in their diagnostics. Model 4 conformed to all of these requirements. 


## c. Visualization

```{r}
# creating new data frame of model predictions for plotting

model_preds_for_plotting <- model_preds %>% 
  rename(water_treatment = x,
         species = group) # renaming columns to make this easier to use

ggplot() +
  # plotting underlying data
  geom_point(data = drought_exp_clean,
             aes(x = water_treatment,
                 y = total_g,
                 color = water_treatment, # sorting by color 
                  alpha = 0.3)) + # making underlying data more transparent
  
  # 95 % CI
    geom_ribbon(data = model_preds_for_plotting,
              aes(x = water_treatment,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high, # setting confidence interval 
                  fill = water_treatment),
              alpha = 0.2) +
  
# model prediction lines
  geom_line(data = model_preds_for_plotting,
            aes(x = water_treatment, 
                y = predicted,
                color = water_treatment)) + # coloring by water treatment  
  
  # cleaner theme
  theme_classic() +
  theme(legend.position = "none") + # removing the legend 
  scale_color_manual(values = c("cornflowerblue", "violet")) + 
# giving each water type a different color 
  
  # creating different panels for species
  facet_wrap(~species_name) +
  
  labs(x = "Water Treatment", # naming x axis 
     y = "Total Mass (g)", # naming y axis
     title = "Total Mass of Species with Different Water Treatments ") # making a title

```

## d. Figure Caption 

**Figure 1:** Jitter plot visualizing the total mass of plants in well watered vs drought stressed water treatments separated by species. Well watered plants were represented with blue while drought stressed plants are represented with pink. Large opaque dots represent the mean total mass (g). Lines represent model prediction lines with 95% confidence intervals. Transparent dots represent underlying data. 

Source: Valliere JM, Zhang J, Sharifi MR, Rundel PW. Can we condition native plants to increase drought tolerance and improve restoration success? Ecol Appl. 2019 Apr;29(3):e01863. doi: 10.1002/eap.1863. Epub 2019 Mar 4. PMID: 30831005.

## e. Results 

My statistical analysis suggests that species and water treatment, or model 4, best influence plant total mass (multiple linear regression model, Normal distribution, F(7, 62) = 27.08, p < 0.001, α = 0.05, R^2 = 0.75). Model 4 also had the lowest AIC (-159.2) and a delta AIC of 0. 

On average across species, well watered plants tend to be 0.12 ± 0.02g heavier than drought stressed plants. Water treatment significantly predicts total mass (p < 0.001).
Across water treatments on average, _Lotus scoparius_ plants were -0.20 ± 0.032 g less than _Encelia californica_ (p < 0.001), and _Grindelia camporum_ plants were -0.31 ± 0.032 less than _Encelia californica_ plants (p < 0.001). 

# Problem 2 

## a.
Because I have time as a variable ranging from 1-10 in my data, I could indicate each minute with a different visual, like a shape or color. This could be separated by another variable like class location, mode of transportation or day of the week. Also, with the variable distance to class, I could also make some sort of map. This is especially relevant because all of the class locations are on campus, which is a relatively small area to draw. 

## b. 
**sketch**

## c. 
**draft**

## d.
In my affective visualization, each colored line indicates an observation in order of time, which is minutes arrived early to class. The key on the left decodes these colors from 1 minute early on the top, to 10 minutes early on the bottom. They go from lightest to darkest. There are five groups separated by class location and labeled on the right hand side. 

I was primarily inspired by Ed Hawkin's "warming stripes" where numerical temperature is represented by a color to showcase a pattern of increasing global temperatures. Similarly, the colors in my piece can help visualize if there is a pattern in arrival time based on class location.

This visual is a poster created in Canva, using their poster format and line tool. Each indivdual line was hand drawn and colored according to the key, and was based on the "minutes early" column in my data. 

# Problem 3

## a. Revisit and summarize 

The author uses a series of independent t tests to assess the impact of a media campaign exposing animal cruelty in the red meat industry on consumer attitudes about the industry before and after watching the campaign. There were 4 data tables that list the before and after eating habits, knowledge/attitudes about farm animal welfare, trust of information sources, and communication in relation to farm animal welfare. 

**Insert the figure or table you described in Homework 2 here.**


## b. Visual clarity 

Summary statistics are clearly listed in each of the 4 data tables. These include the t value, df, p-value, means (pre, post and difference) and standard error. However the authors did not visually represent their data in this paper, making it harder to comprehend the results or see the underlying data. There is also no clear way to access the data in this paper, meaning the summary statistics are the only way to analyze the results as readers. 

 
## c. Aesthetic clarity

With 4 separate data tables describing the results of the survey, there is a lot of information to navigate to find the most relevant data in relation to the central question. There is not only a lot of "clutter" within some of the tables with unclear topics, but the tables themselves sometimes feel cluttered as well. For example, table 6, which describes respondents beliefs about animal welfare before and after the program contains vague and confusing topics and variables that are not explained in the paper. These include topics of "perceived knowledge", "actual knowledge", and "knowledge score' with no explanation to how these were calculated. However, the data tables are consistently formatted with their summary statistics making it easier to understand the information presented in each table. 

## d. Recommendations 

A jitter plot would be a good way to visualize the data in this paper as it can compare two groups of means for one variable, being the before and after survey responses for each  question. The x-axis would a variable, and the y axis would be the 1-10 scale response from each survey participant. The before and after could be different colors (with a legend) to see if there is overlap in these points. There should also be opaque dots signifying the mean with error bars, and the underlying data should be more transparent.  

Furthermore, as mentioned earlier, there should be an explanation of how some of the variables are measured, such as "perceived knowledge" vs "actual knowledge". It is unclear as readers to how these means were calculated (via a test, or self reported?). It would also be helpful if there was an example of the survey to understand how these questions were specifically worded to improve readers understanding of variables listed. 
