---
title: "Homework 3"
author: "Kirya Wagner"
GitHub Link: "https://github.com/kiryawagner/Wagner-Kirya-Homework-03"
format:
  html: 
    toc: true # includes a table of contents in rendered html format
execute: # for all code chunks
    warning: false # do not display warnings
    message: false # do not display messages
---

**WARNING:** this document provides the foundation for your homework 3, but you will need to create your own neat, organized document to render for your homework submission. Create a new document for your homework in the `code` folder of this repository.   

**You will not need all the information in this document to do your homework; IT IS YOUR JOB TO EDIT THE INFORMATION IN YOUR HOMEWORK DOCUMENT TO ONLY INCLUDE WHAT IS NECESSARY.**    

**Relevant definitions:**  

specific leaf area (SLA, units: mm^2^/g): surface area of a leaf divided by its dry mass, represents life strategy (high SLA = large surface area and low mass = "fast" species)  

root mass (units: g): mass of root tissue  

shoot mass (units: g): mass of aboveground tissue  

# Set up

## reading in packages

```{r}
# general use
library(tidyverse)
library(readxl)
library(here)
library(janitor)

# visualizing pairs
library(GGally)

# model selection
library(MuMIn)

# model predictions
library(ggeffects)

# model tables
library(gtsummary)
library(flextable)
library(modelsummary)

drought_exp <- read_xlsx(path = here("data", 
                                     "Valliere_etal_EcoApps_Data.xlsx"),
                         sheet = "First Harvest")

# here allows us to refer to specific. first arugment data file, then whawt the file actually is. then sheet

# quick look at data 
str(drought_exp) # gives a tibble of what is in each column
class(drought_exp) # tells us what kind of "thing" each object is. Ex: table and data frame 
```

## cleaning

```{r}
# cleaning
drought_exp_clean <- drought_exp %>% 
  clean_names() %>% # nicer column names
  mutate(species_name = case_when( # adding column with species scientific names
    species == "ENCCAL" ~ "Encelia californica", # bush sunflower
    species == "ESCCAL" ~ "Eschscholzia californica", # California poppy
    species == "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    species == "GRICAM" ~ "Grindelia camporum", # great valley gumweed
    species == "SALLEU" ~ "Salvia leucophylla", # Purple sage
    species == "STIPUL" ~ "Nasella pulchra", # Purple needlegrass
    species == "LOTSCO" ~ "Acmispon glaber" # deerweed
  )) %>% 
  relocate(species_name, .after = species) %>% # moving species_name column after species
  mutate(water_treatment = case_when( # adding column with full treatment names
    water == "WW" ~ "Well watered",
    water == "DS" ~ "Drought stressed"
  )) %>% 
  relocate(water_treatment, .after = water) # moving water_treatment column after water
```

# Visualizing

## correlations 

```{r}
ggpairs(drought_exp_clean, # data frame
        columns = c("leaf_dry_weight_g", # columns to visualize
                    "sla", 
                    "shoot_g", 
                    "root_g", 
                    "total_g"), 
        upper = list(method = "pearson")) + # calculating Pearson correlation coefficient
  theme_bw() + # cleaner theme
  theme(panel.grid = element_blank()) # getting rid of gridlines

# visualize and calculate correlation coefficient
# ex: top left dist of leaf dry weight (smooth diagonal line is distribution of all variable)
# bottom left: scatter plot (10 eom point style plots for all of continious variables). EX: bottom left relationshap between leaf dry weight and total mass  
# root mass, leaf dry weight corr = 0.367. signficantly/strongly positively correlated 
```

Choose 3 pairs of variables to compare. Make notes of those comparisons below.    

**Leaf dry weight is weakly and positively correlated with specific leaf area (Pearson's r = 0.033**  

**Shoot mass and total mass are strongly positively and correlated (Pearson's r = 0.902)**  

**Leaf dry weight is strongly and positively correlated with total mass (Pearson's r = 0.046**

## Preliminary visualizations  

If you aren't familiar with each plant species, google them!  

Then go exploring! Visually.  

In this scenario, we're interested in how three variables affect total biomass: species (categorical), water treatment (categorical), specific leaf area (SLA, numeric).  

**Note:** it is totally ok (and encouraged) to explore these separately!

### how does total biomass differ between species?

```{r}
ggplot(data = drought_exp_clean, # data frame
       aes(x = reorder(species_name, # reordering x-axis
                       -total_g, # in reverse order of mean total mass
                       fun = mean), # calculating mean to reorder
           y = total_g)) + # y-axis
  geom_jitter(width = 0.1, # narrow jitter
              height = 0) # not jittering points up and down

# one easy way to get out patterns is to reorder x axis in reverse order of mean total mass 

```

**Total biomass seemingly does differ between species. Gumweed, California poppy and purple needlegrass seem to have the highest biomass. Deerweed, scarlet buguler and purplse sage seem to have the lowest biomass.**



### how does total biomass differ between water treatments?


```{r}
ggplot(data = drought_exp_clean, # data frame
       aes(x = water_treatment, # x-axis
           y = total_g)) + # y-axis
  geom_jitter(width = 0.1, # narrow jitter
              height = 0) # not jittering points up and down
```

**It seems that well water treated plants have a higher total biomass than drought stressed plants.**

### how does specific leaf area (SLA) influence total biomass?

```{r}
ggplot(data = drought_exp_clean, # data frame
       aes(x = sla, # x-axis
           y = total_g)) + # y-axis
  geom_point() # scatterplot
```

**It seems that specific leaf area moderately and positively affects total leaf mass.**

# Model construction

What is the null model?  

0. **Total biomass as a function of none of the predictors**

What is the saturated model?  

1. **Total biomass as a function of Specific leaf area, water treatment and species**  

In class we are going to try two more models:  

2.  **total biomass as a function of SLA and water treatment**  
3.  **total biomass as a function of SLA and species**  

For homework, you will construct one more model:  

4. **total biomass as a function of water treatment and species**  

You will then compare all 5 models to determine which best describes total biomass. **For your homework, you will need to create a table summarizing the structure of these models.**

## 0. Null model

```{r}
model0 <- lm(total_g ~ 1, # formula
             data = drought_exp_clean) # data frame
```

## 1. total biomass as a function of SLA, water treatment, and species

```{r}
# saturated model
model1 <- lm(total_g ~ sla + water_treatment + species_name,
             data = drought_exp_clean)

par(mfrow = c(2, 2))
plot(model1)
# you might get a warning when you run this code - that is ok!
```
**Diagnostics looks ok!**

## 2. total biomass as a function of SLA and water treatment

```{r}
model2 <- lm(total_g ~ sla + water_treatment,
             data = drought_exp_clean)

plot(model2)
```

**LOOK AT DIAGNOSTICS**
**Diagnostics look good**

## 3. total biomass as a function of SLA and species

```{r}
model3 <- lm(total_g ~ sla + species_name,
             data = drought_exp_clean)

plot(model3)
```

**LOOK AT DIAGNOSTICS**
**The QQ plot slightly derails from the line at the tails, but the diagnostics overall look okay. 

## 4. total biomass as a function of water treatment and species

```{r}
model4 <- lm(total_g ~ water_treatment + species_name,
             data = drought_exp_clean)

plot(model3)
```

**LOOK AT DIAGNOSTICS**
Diagnostics overall look okay. 

# Model selection

From the `MuMIn` package

```{r}
# compare functions 
# "multi model inference"
model.sel(model0,
          model1, 
          model2, 
          model3,
          model4)
```

Out of the four models we have tried, the best model is **model 4, the one that includes biomass as a function water treatment and species**. This is because its delta is 0, and it has the lowest AIC. 

**BEST model (balances explanatory power with complecity) is model 1, or the saturated model. Delta: difference between AIC of best and others. Always 0 for best model.**  
**DOUBLE CHECK DIAGNOSTICS FOR BEST MODEL**

Note that this may change once you fit and compare the other model!

# Model summary

```{r}
summary(model1)
```

What are the reference levels?  

- reference for `water_treatment`:
- reference for `species_name`: 

**reference for water_treatment:** **drought stress**  
**reference for species_name:** **acmispon glaber (deerweed)**


Interpret the model estimates for the:  

- intercept: **mass of the reference (Acmispon glaber) in drought stressed water treatment when SLA = 0**  
- SLA slope: **for each 1 unit incresase in SLA (mm^2^/g), you expect a decrease in plant mass of 0.002 +- (figure out how to display plus minus in doc) 0.005. NO EFFECT OF SLA ON MASS EVEN THOUGH CHOSEN AS BEST MODEL**  
- well watered estimate: **on average across species and SLA, well watered plants tend to be 0.12 +- 0.02g heavier than drought stressed plants. Water treatment signficantly predicts total mass (with p)**    

- Lotus scoparius estimate: **insert response here**    
- Stipa pulchra estimate: **insert response here**    

# Model predictions

**Note:** only plot terms in the model you select - if your doesn't include one of these terms, take it out and adjust the plotting code accordingly!

```{r}
model_preds <- ggpredict(model1, 
                         terms = c("sla",
                                   "water_treatment", 
                                   "species_name"))

view(model_preds)

# use View(model_preds) to see the predictions as a data frame
# use model_preds to see the predictions formatted nicely
```

# Visualization

**Note:** this is the quick way to visualize model predictions. For your homework, you may find it easier to plot things yourself (i.e. not using the `plot()` functionality within `ggeffects`). See lecture for an example of how to do this.  

```{r}
# plot function within ggeffects
plot(model_preds, # model predictions
     limit_range = TRUE, # limit the range of predictions to the range of predictor values
     show_data = TRUE) + # show the underlying data
  # everything below this is ggplot() stuff
  theme_classic() + # classic theme
  labs(title = "Preliminary model visualization") + # plot title
  theme(panel.grid = element_blank()) # getting rid of gridlines
```

The "more customizable" way:

```{r}
# creating new data frame of model predictions for plotting
model_preds_for_plotting <- model_preds %>% 
  rename(sla = x, # renaming columns to make this easier to use
         water_treatment = group,
         species_name = facet)

# use View(model_preds_for_plotting) 
# to compare this to the original model_preds data frame

ggplot() +
  # underlying data
  geom_point(data = drought_exp_clean,
             aes(x = sla,
                 y = total_g,
                 color = water_treatment)) +
  # model prediction 95% CI ribbon
  geom_ribbon(data = model_preds_for_plotting,
              aes(x = sla, 
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high,
                  fill = water_treatment),
              alpha = 0.2) +
  # model prediction lines
  geom_line(data = model_preds_for_plotting,
            aes(x = sla, 
                y = predicted,
                color = water_treatment)) +
  # cleaner theme
  theme_classic() +
  # creating different panels for species
  facet_wrap(~species_name) 
```


# Making a table

```{r}
# one option for a single model
flextable::as_flextable(model1) 

# another option for a single model
gtsummary::tbl_regression(model1)

# comparing models
modelsummary::modelsummary( # this function takes a list of models
  list( 
    "null" = model0, # "model name" = model object
    "model 1" = model1,
    "model 2" = model2,
    "model 3" = model3
  )
) 

```

**Note 1: notice all the information at the bottom of the `modelsummary` output - if you only needed the AIC and delta AIC, what could you do? see package documentation for help**

**Note 2: you will always have to report the F-statistic, degrees of freedom, test statistic, p-value, ⍺, and R^2^. Whether or not this information is in a table is up to you.**

