---
title: "Homework 3"
author: "Kirya Wagner"
GitHub Link: "https://github.com/kiryawagner/Wagner-Kirya-Homework-03"
format:
  html: 
    toc: true # includes a table of contents in rendered html format
execute: # for all code chunks
    warning: false # do not display warnings
    message: false # do not display messages
---

**WARNING:** this document provides the foundation for your homework 3, but you will need to create your own neat, organized document to render for your homework submission. Create a new document for your homework in the `code` folder of this repository.   

**You will not need all the information in this document to do your homework; IT IS YOUR JOB TO EDIT THE INFORMATION IN YOUR HOMEWORK DOCUMENT TO ONLY INCLUDE WHAT IS NECESSARY.**    

**Relevant definitions:**  

specific leaf area (SLA, units: mm^2^/g): surface area of a leaf divided by its dry mass, represents life strategy (high SLA = large surface area and low mass = "fast" species)  

root mass (units: g): mass of root tissue  

shoot mass (units: g): mass of aboveground tissue  

# Set up

## reading in packages

```{r}
# general use
library(tidyverse)
library(readxl)
library(here)
library(janitor)

# visualizing pairs
library(GGally)

# model selection
library(MuMIn)

# model predictions
library(ggeffects)

# model tables
library(gtsummary)
library(flextable)
library(modelsummary)

drought_exp <- read_xlsx(path = here("data", 
                                     "Valliere_etal_EcoApps_Data.xlsx"),
                         sheet = "First Harvest")

# here allows us to refer to specific. first arugment data file, then whawt the file actually is. then sheet

# quick look at data 
str(drought_exp) # gives a tibble of what is in each column
class(drought_exp) # tells us what kind of "thing" each object is. Ex: table and data frame 
```

## cleaning

```{r}
# cleaning
drought_exp_clean <- drought_exp %>% 
  clean_names() %>% # nicer column names
  mutate(species_name = case_when( # adding column with species scientific names
    species == "ENCCAL" ~ "Encelia californica", # bush sunflower
    species == "ESCCAL" ~ "Eschscholzia californica", # California poppy
    species == "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    species == "GRICAM" ~ "Grindelia camporum", # great valley gumweed
    species == "SALLEU" ~ "Salvia leucophylla", # Purple sage
    species == "STIPUL" ~ "Nasella pulchra", # Purple needlegrass
    species == "LOTSCO" ~ "Acmispon glaber" # deerweed
  )) %>% 
  relocate(species_name, .after = species) %>% # moving species_name column after species
  mutate(water_treatment = case_when( # adding column with full treatment names
    water == "WW" ~ "Well watered",
    water == "DS" ~ "Drought stressed"
  )) %>% 
  relocate(water_treatment, .after = water) # moving water_treatment column after water
```

# Visualizing

## correlations 

```{r}
ggpairs(drought_exp_clean, # data frame
        columns = c("leaf_dry_weight_g", # columns to visualize
                    "sla", 
                    "shoot_g", 
                    "root_g", 
                    "total_g"), 
        upper = list(method = "pearson")) + # calculating Pearson correlation coefficient
  theme_bw() + # cleaner theme
  theme(panel.grid = element_blank()) # getting rid of gridlines

# visualize and calculate correlation coefficient
# ex: top left dist of leaf dry weight (smooth diagonal line is distribution of all variable)
# bottom left: scatter plot (10 eom point style plots for all of continious variables). EX: bottom left relationshap between leaf dry weight and total mass  
# root mass, leaf dry weight corr = 0.367. signficantly/strongly positively correlated 
```

Choose 3 pairs of variables to compare. Make notes of those comparisons below.    

**Leaf dry weight is weakly and positively correlated with specific leaf area (Pearson's r = 0.033**  

**Shoot mass and total mass are strongly positively and correlated (Pearson's r = 0.902)**  

**Leaf dry weight is strongly and positively correlated with total mass (Pearson's r = 0.046**

## Preliminary visualizations  

If you aren't familiar with each plant species, google them!  

Then go exploring! Visually.  

In this scenario, we're interested in how three variables affect total biomass: species (categorical), water treatment (categorical), specific leaf area (SLA, numeric).  

**Note:** it is totally ok (and encouraged) to explore these separately!

### how does total biomass differ between species?

```{r}
ggplot(data = drought_exp_clean, # data frame
       aes(x = reorder(species_name, # reordering x-axis
                       -total_g, # in reverse order of mean total mass
                       fun = mean), # calculating mean to reorder
           y = total_g)) + # y-axis
  geom_jitter(width = 0.1, # narrow jitter
              height = 0) # not jittering points up and down

# one easy way to get out patterns is to reorder x axis in reverse order of mean total mass 

```

**Total biomass seemingly does differ between species. Gumweed, California poppy and purple needlegrass seem to have the highest biomass. Deerweed, scarlet buguler and purplse sage seem to have the lowest biomass.**



### how does total biomass differ between water treatments?


```{r}
ggplot(data = drought_exp_clean, # data frame
       aes(x = water_treatment, # x-axis
           y = total_g)) + # y-axis
  geom_jitter(width = 0.1, # narrow jitter
              height = 0) # not jittering points up and down
```

**It seems that well water treated plants have a higher total biomass than drought stressed plants.**

### how does specific leaf area (SLA) influence total biomass?

```{r}
ggplot(data = drought_exp_clean, # data frame
       aes(x = sla, # x-axis
           y = total_g)) + # y-axis
  geom_point() # scatterplot
```

**It seems that specific leaf area moderately and positively affects total leaf mass.**

# Model construction

What is the null model?  

0. **Total biomass as a function of none of the predictors**

What is the saturated model?  

1. **Total biomass as a function of Specific leaf area, water treatment and species**  

In class we are going to try two more models:  

2.  **total biomass as a function of SLA and water treatment**  
3.  **total biomass as a function of SLA and species**  

For homework, you will construct one more model:  

4. **total biomass as a function of water treatment and species**  

You will then compare all 5 models to determine which best describes total biomass. **For your homework, you will need to create a table summarizing the structure of these models.**

## 0. Null model

```{r}
model0 <- lm(total_g ~ 1, # formula
             data = drought_exp_clean) # data frame
```

## 1. total biomass as a function of SLA, water treatment, and species

```{r}
# saturated model
model1 <- lm(total_g ~ sla + water_treatment + species_name,
             data = drought_exp_clean)

par(mfrow = c(2, 2))
plot(model1)
# you might get a warning when you run this code - that is ok!
```
**Diagnostics looks ok!**

## 2. total biomass as a function of SLA and water treatment

```{r}
model2 <- lm(total_g ~ sla + water_treatment,
             data = drought_exp_clean)

plot(model2)
```

**LOOK AT DIAGNOSTICS**
**Diagnostics look good**

## 3. total biomass as a function of SLA and species

```{r}
model3 <- lm(total_g ~ sla + species_name,
             data = drought_exp_clean)

plot(model3)
```

**LOOK AT DIAGNOSTICS**
**The QQ plot slightly derails from the line at the tails, but the diagnostics overall look okay. 

## 4. total biomass as a function of water treatment and species

```{r}
model4 <- lm(total_g ~ water_treatment + species_name,
             data = drought_exp_clean)

plot(model4)
```

**LOOK AT DIAGNOSTICS**
Diagnostics overall look okay. 

# Model selection



```{r}
# compare functions 

model.sel(model0,
          model1, 
          model2, 
          model3,
          model4)
```

Out of the four models we have tried, the best model is **model 4, the one that includes biomass as a function water treatment and species**. This is because its delta is 0, and it has the lowest AIC. 

**DOUBLE CHECK DIAGNOSTICS FOR BEST MODEL**

Note that this may change once you fit and compare the other model!

# Model summary

```{r}
summary(model4)
```
- reference for `water_treatment`: **drought stress**  
- reference for `species_name`: **acmispon glaber (deerweed)**

- intercept: **The intercept, or mass of the reference (Acmispon glaber) in drought stressed water treatment, is 0.05455 ± 0.024g**  

- well watered estimate: **on average across species, well watered plants tend to be 0.12 ± 0.02g heavier than drought stressed plants. Water treatment significantly predicts total mass as p < 0.001.**    

**_Lotus scoparius_ estimate:** across water treatments on average, _Lotus scoparius_ plants were -0.20 ± 0.032 g less than _Encelia californica_.  

**_Stipa pulchra_ estimate:** across water treatments, _Stipa pulchra_ plants are roughly the same size as _Encelia californica plants_.


# Model predictions

**Note:** only plot terms in the model you select - if your doesn't include one of these terms, take it out and adjust the plotting code accordingly!

```{r}
model_preds <- ggpredict(model4, 
                         terms = c("water_treatment", 
                                   "species_name"))

view(model_preds)

# use View(model_preds) to see the predictions as a data frame
# use model_preds to see the predictions formatted nicely
```



# Making a table

Table 1: Table displaying model estimates for the null and models 1-4.

```{r}

# comparing models
modelsummary::modelsummary # this function takes a list of models
  list( 
    "null" = model0, # "model name" = model object
    "model 1" = model1,
    "model 2" = model2,
    "model 3" = model3,
    "model 4" = model4)

```

**Note 1: notice all the information at the bottom of the `modelsummary` output - if you only needed the AIC and delta AIC, what could you do? see package documentation for help**

**Note 2: you will always have to report the F-statistic, degrees of freedom, test statistic, p-value, ⍺, and R^2^. Whether or not this information is in a table is up to you.**


## statistical methods 

In this statistical analysis, I examined the influence of specific leaf area (SLA), species, and water treatment on total growth (g) of the plant by constructing a set of 5 models that compared each variable. 

These included total biomass as a function of none of the predictors, total biomass as a function of SLA, water treatment and species, total biomass as a function of SLA and water treatment, total biomass as a function of SLA and species, and total biomass as a function of water treatment and species. 

First, I visualized the variables individually against total mass and determined that there seemingly was a correlation between each one. Then I created the models using lm(), and looked at the residuals, distributions, and outlines within their diagnostics to determine if the model was "valid", or "usable". 
Next, I compared each model using model.sel(), and determined which one was the best based on their diagnostics and coefficients. The best model not only has a lower AIC and delta of 0, but also visually needs homeostatic residuals, a normal distribution and no influential outliers in their diagnostics. Model 4 met all of these requirements 


# Visualization

```{r model-predictions-with-ggeffects}
# creating new data frame of model predictions for plotting
model_preds_for_plotting <- model_preds %>% 
  rename(water_treatment = x,
         species = group) # renaming columns to make this easier to use

ggplot() +
  # underlying data
  geom_point(data = drought_exp_clean,
             aes(x = water_treatment,
                 y = total_g,
                 color = water_treatment,
                  alpha = 0.3)) +

  # model prediction lines
  geom_line(data = model_preds_for_plotting,
            aes(x = water_treatment, 
                y = predicted,
                color = water_treatment)) +
  
  # cleaner theme
  theme_classic() +
  theme(legend.position = "none") +
  scale_color_manual(values = c("cornflowerblue", "violet")) + # giving each water type a different color 
  
  # creating different panels for species
  facet_wrap(~species_name) +
  
  labs(x = "Water Treatment", # naming x axis 
     y = "Total Mass (g)", # naming y axis
     title = "Total Mass of Species with Different Water Treatments ") # making a title

```
Figure 1: Jitter plot visualizing the total mass of plant species in well watered vs drought stressed water treatments. Well watered plants were represented with blue while drought stressed plants are represented with pink. Large opague dots represent the mean total mass (g). Lines represent model predition lines. Trasnparent dots represent underlying data. 

Source: Valliere JM, Zhang J, Sharifi MR, Rundel PW. Can we condition native plants to increase drought tolerance and improve restoration success? Ecol Appl. 2019 Apr;29(3):e01863. doi: 10.1002/eap.1863. Epub 2019 Mar 4. PMID: 30831005.


